Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-07-25 submission/manuscript.Rmd
put down new results section

**Decreased SCFA concentrations are not associated with tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time. There was no difference between individuals without colon tumors (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F]. Even though there was no difference in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.

@@

**The number of significant positive correlations between OTU relative abundance and SCFA concentration are similar for individuals without tumors and patients with adenoma or carcinoma tumors.** Having found no difference in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. Using Spearmans rho we found that the majority of OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5]. A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6]. There was a noticeablely higher number of significant negative correlations associated with patients with adenoma versus the other groups for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family seemed to have the largest share of these negative correlations within patients with adenoma tumors [Figure 3]. The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3]. Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to the other groups than the number of negative correlations [Figure 3]. Collectively, these results suggest that the resident taxa that may change the most due to carcinoma tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

**Random Forest models with SCFA concentrations do not classify tumors better.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures. Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors. Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4]. Although including SCFA concentrations did not improve classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor. This would be consistent with previous 16S rRNA gene sequencing results where many taxa associated with SCFA production are decreased [@Shah2017; @meta_analysis_crc_Sze2018].

**Expected taxa are associated with higher SCFA concentrations regardless of tumor status.** Using OTU data we built Random Forest models to classify higher than median and lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A]. However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [Figure S1A]. The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were taxa that are normally associated with SCFA production [Figure S1B]. These results highlight that within our data, SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate. Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations. Overall, our results are robust and do not support the hypothesis that differences in SCFA concentrations are associated with colorectal tumors.

>>>>

**Decreased SCFA concentrations are not associated with tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time. There was no difference between individuals without colon tumors (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F].

@@

**The number of significant positive correlations between OTU relative abundance and SCFA concentration are similar for individuals without tumors and patients with adenoma or carcinoma tumors.** Having found no difference in SCFA concentrations or genes coding for enzymes involved with SCFA synthesis, we next investigated if specific OTUs correlated with SCFA concentrations. Using Spearmans rho we found that the majority of OTU correlations were to taxa from *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* [Figure 3 & Table S5]. A similar pattern was observed when using high/low SCFA groups based on the overall median concentration for that specific SCFA [Figure S1 & Table S6]. There was a noticeablely higher number of significant negative correlations associated with patients with adenoma versus the other groups for all SCFAs tested [Figure 3]. In particular, OTUs from the *Ruminococcaceae* family seemed to have the largest share of these negative correlations within patients with adenoma tumors [Figure 3]. The number of positive correlations between OTUs and SCFA concentrations was similar between individuals without tumors and patients with a carcinoma [Figure 3]. Although patients with adenomas had more positive correlations between OTUs and SCFA concentrations, their total number was more similar to the other groups than the number of negative correlations [Figure 3]. Overall, these results suggest that the resident taxa that may change the most due to carcinoma tumors may not be ones that are responsible for the production of acetate, butyrate, or propionate.

**SCFA concentrations do not replace all important *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* OTUs in Random Forest models built to classify tumors.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures. Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors. Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4A & 4D]. There was little difference between the top 10 most important OTUs, as measured by mean decrease in accuracy, in models using SCFA and OTUs versus OTUs only [Figure 4B-C & 4E-F]. The reason for this seems to be because SCFA concentrations replaced the information gained by specific OTUs rather than add new information to the model [Figure 4B-C]. The SCFA concentrations also only replaced select OTUs with taxonomic classification to *Clostridiales*, *Lachnospiraceae*, and *Ruminococcaceae* with many OTUs within these taxa remaining [Figure 4B & 4C]. In combination with the previous results on taxa correlations, these observations provide additional evidence that the resident taxa within models that classify tumor are not ones associated with acetate, butyrate, or propionate production.

**Random Forest models for SCFA concentrations have different top 10 most important OTUs than disease models.** Using OTU data we built Random Forest models to classify either SCFA concentration or higher than median and lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify both SCFA concentrations and high and low SCFA concentrations [Figure 5A & S2A]. However, these models tended to be over fit, suggesting that rarer taxa may be important for this classification [[Figure 5A & S2A]. Additionally, there were differences classification for both models based on whether the individual had no tumors, an adenoma, or a carcinoma [Figure 5B & S2B]. There was minimal overlap between these model's most important OTUs and those used to classify patients with either adenoma or carcinoma tumors [Figure 4B-C, 4E-F, 5C-E, and S2C-E]. The only OTUs that did overlap between the models was OTU00167 (*Clostridiales*) [Figure 4B-C, 4E-F, 5C-E, and S2C-E]. OTU00167 was in the top 10 most important OTUs for the OTU adenoma model and not in the SCFA and OTU adenoma model, however, acetate and butyrate concentrations were [Figure 4B-C]. These observations provide further evidence that it is possible to identify specific OTUs associated with higher SCFA concentrations and that these OTUs belong to taxa known to produce acetate, butyrate, and propionate. Although it is possible to identify OTUs associated with SCFA production, our results do not support the hypothesis that SCFA concentration or OTUs associated with their production are different between individuals with no tumors and patients with an adenoma or carcinoma tumor.

### 2018-07-25 submission/manuscript.tex
put down new results section

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used high-performance liquid chromatography (HPLC) to measure
acetate, butyrate, and propionate concentrations of frozen fecal samples
from 490 individuals at a cross-sectional point in time. There was no
difference between individuals without colon tumors (n=172) and patients
with either an adenoma (n=198) or carcinoma (n=120) for any of the SCFAs
measured after multiple comparison correction (P-value \textgreater{}
0.15) {[}Figure 1A - 1C{]}. We next measured the concentration of SCFAs
in 67 patients with an adenoma (n=41) or carcinoma (n=26) in which we
had pre- and post-treatment fecal samples. Although there was a general
trend for increasing acetate, butyrate, and propionate concentrations
after treatment for tumors, there was no significant difference pre- and
post-treatment for patients with adenoma (P-value \textgreater{} 0.20)
or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even
though there was no difference in SCFA concentrations between
individuals with normal colons and those with tumors, this information
could still be important to help classify disease.

\textbf{Random Forest models with SCFA concentrations do not classify
tumors better.} SCFA concentrations could improve prediction of tumors
based on specific bacterial community structures. Our OTU data can be
used in combination with SCFAs to assess whether there is a community
dependent context to SCFA classification of tumors. Using the Random
Forest algorithm we built models with OTU abundance data only or OTU
abundances and SCFA concentrations to classify normal versus adenoma and
normal versus carcinoma fecal samples. For adenoma and carcinoma models,
there was no difference between the median AUC of models with or without
SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
Although including SCFA concentrations did not improve classification of
colorectal tumors using Random Forest models, it is possible that the
genes for enzymes involved in SCFA synthesis may vary based on the type
of colorectal tumor. This would be consistent with previous 16S rRNA
gene sequencing results where many taxa associated with SCFA production
are decreased (10, 11).

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance between individuals without colon tumors
and patients with adenomas or carcinomas. First, using imputed gene
relative abundance based on 16S rRNA gene sequencing we found no
difference in any of the genes involved with acetate, butyrate, or
propionate synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. This
similarity between groups is highlighted by visualizing genes important
in butyrate synthesis {[}Figure 3A{]}. Using a paired Wilcoxon rank-sum
test, there also was no difference in imputed gene relative abundance
between pre- and post-treatment samples for any genes involved with SCFA
synthesis (P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a
subset of these 490 fecal samples (n=85) and used metagenomic sequencing
to confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals without colon tumors (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference is highlighted when we
visualize the results for butyrate kinase {[}Figure 3B{]}. These
observations provide evidence that gene prevalence does not change due
to colorectal tumors.

\textbf{Expected taxa are associated with higher SCFA concentrations
regardless of tumor status.} Using OTU data we built Random Forest
models to classify higher than median and lower than median SCFA
concentrations. Overall, OTU data had a reasonable ability to classify
high and low SCFA concentrations {[}Figure S1A{]}. However, these models
tended to be over fit, suggesting that rarer taxa may be important for
this classification {[}Figure S1A{]}. The most important OTUs to these
models (assessed with mean decrease in accuracy (MDA)) were taxa that
are normally associated with SCFA production {[}Figure S1B{]}. These
results highlight that within our data, SCFA concentrations are
associated with taxa known to produce acetate, butyrate, and propionate.
Additionally, OTUs associated with these taxa are the most important to
models that can classify high and low SCFA concentrations. Overall, our
results are robust and do not support the hypothesis that differences in
SCFA concentrations are associated with colorectal tumors.

>>>>

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used high-performance liquid chromatography (HPLC) to measure
acetate, butyrate, and propionate concentrations of frozen fecal samples
from 490 individuals at a cross-sectional point in time. There was no
difference between individuals without colon tumors (n=172) and patients
with either an adenoma (n=198) or carcinoma (n=120) for any of the SCFAs
measured after multiple comparison correction (P-value \textgreater{}
0.15) {[}Figure 1A - 1C{]}. We next measured the concentration of SCFAs
in 67 patients with an adenoma (n=41) or carcinoma (n=26) in which we
had pre- and post-treatment fecal samples. Although there was a general
trend for increasing acetate, butyrate, and propionate concentrations
after treatment for tumors, there was no significant difference pre- and
post-treatment for patients with adenoma (P-value \textgreater{} 0.20)
or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance between individuals without colon tumors
and patients with adenomas or carcinomas. First, using imputed gene
relative abundance based on 16S rRNA gene sequencing we found no
difference in any of the genes involved with acetate, butyrate, or
propionate synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. This
similarity between groups is highlighted by visualizing genes important
in butyrate synthesis {[}Figure 2A{]}. Using a paired Wilcoxon rank-sum
test, there also was no difference in imputed gene relative abundance
between pre- and post-treatment samples for any genes involved with SCFA
synthesis (P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a
subset of these 490 fecal samples (n=85) and used metagenomic sequencing
to confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals without colon tumors (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference is highlighted when we
visualize the results for butyrate kinase {[}Figure 2B{]}. These
observations provide evidence that gene prevalence does not change due
to colorectal tumors.

\textbf{The number of significant positive correlations between OTU
relative abundance and SCFA concentration are similar for individuals
without tumors and patients with adenoma or carcinoma tumors.} Having
found no difference in SCFA concentrations or genes coding for enzymes
involved with SCFA synthesis, we next investigated if specific OTUs
correlated with SCFA concentrations. Using Spearmans rho we found that
the majority of OTU correlations were to taxa from \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} {[}Figure 3 \& Table
S5{]}. A similar pattern was observed when using high/low SCFA groups
based on the overall median concentration for that specific SCFA
{[}Figure S1 \& Table S6{]}. There was a noticeablely higher number of
significant negative correlations associated with patients with adenoma
versus the other groups for all SCFAs tested {[}Figure 3{]}. In
particular, OTUs from the \emph{Ruminococcaceae} family seemed to have
the largest share of these negative correlations within patients with
adenoma tumors {[}Figure 3{]}. The number of positive correlations
between OTUs and SCFA concentrations was similar between individuals
without tumors and patients with a carcinoma {[}Figure 3{]}. Although
patients with adenomas had more positive correlations between OTUs and
SCFA concentrations, their total number was more similar to the other
groups than the number of negative correlations {[}Figure 3{]}. Overall,
these results suggest that the resident taxa that may change the most
due to carcinoma tumors may not be ones that are responsible for the
production of acetate, butyrate, or propionate.

\textbf{SCFA concentrations do not replace all important
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
OTUs in Random Forest models built to classify tumors.} SCFA
concentrations could improve prediction of tumors based on specific
bacterial community structures. Our OTU data can be used in combination
with SCFAs to assess whether there is a community dependent context to
SCFA classification of tumors. Using the Random Forest algorithm we
built models with OTU abundance data only or OTU abundances and SCFA
concentrations to classify normal versus adenoma and normal versus
carcinoma fecal samples. For adenoma and carcinoma models, there was no
difference between the median AUC of models with or without SCFA
concentrations (P-value \textgreater{} 0.05) {[}Figure 4A \& 4D{]}.
There was little difference between the top 10 most important OTUs, as
measured by mean decrease in accuracy, in models using SCFA and OTUs
versus OTUs only {[}Figure 4B-C \& 4E-F{]}. The reason for this seems to
be because SCFA concentrations replaced the information gained by
specific OTUs rather than add new information to the model {[}Figure
4B-C{]}. The SCFA concentrations also only replaced select OTUs with
taxonomic classification to \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} with many OTUs within
these taxa remaining {[}Figure 4B \& 4C{]}. In combination with the
previous results on taxa correlations, these observations provide
additional evidence that the resident taxa within models that classify
tumor are not ones associated with acetate, butyrate, or propionate
production.

\textbf{Random Forest models for SCFA concentrations have different top
10 most important OTUs than disease models.} Using OTU data we built
Random Forest models to classify either SCFA concentration or higher
than median and lower than median SCFA concentrations. Overall, OTU data
had a reasonable ability to classify both SCFA concentrations and high
and low SCFA concentrations {[}Figure 5A \& S2A{]}. However, these
models tended to be over fit, suggesting that rarer taxa may be
important for this classification {[}{[}Figure 5A \& S2A{]}.
Additionally, there were differences classification for both models
based on whether the individual had no tumors, an adenoma, or a
carcinoma {[}Figure 5B \& S2B{]}. There was minimal overlap between
these model's most important OTUs and those used to classify patients
with either adenoma or carcinoma tumors {[}Figure 4B-C, 4E-F, 5C-E, and
S2C-E{]}. The only OTUs that did overlap between the models was OTU00167
(\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E, and S2C-E{]}.
OTU00167 was in the top 10 most important OTUs for the OTU adenoma model
and not in the SCFA and OTU adenoma model, however, acetate and butyrate
concentrations were {[}Figure 4B-C{]}. These observations provide
further evidence that it is possible to identify specific OTUs
associated with higher SCFA concentrations and that these OTUs belong to
taxa known to produce acetate, butyrate, and propionate. Although it is
possible to identify OTUs associated with SCFA production, our results
do not support the hypothesis that SCFA concentration or OTUs associated
with their production are different between individuals with no tumors
and patients with an adenoma or carcinoma tumor.
